\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-04-25 doc/incertae_sedis.tex
Tweak wording in 1.2 and move figure.

The problem is becoming more acute because NCBI is putting a larger
number of taxa into groups that are marked as “unclassified.” For
example, when OTT 2.9 was created NCBI’s clasification of the bird
family Sylvidae included a group five genera that were placed within
“[http://www.ncbi.nlm.nih.gov/Taxonomy/Brows er/wwwtax.cgi?mode=Undef&
id=36270&lvl=3&lin=f&keep=1&srchmode=1&unlock||unclassi fied
Sylviidae].” This includes the genus Regulus.
Thus in OTT 2.9 Regulus
(ott ID = 3599326) is placed inside the family Silvidae (OTT ID =
259942).
The “unclassified Sylviidae” does not appear in OTT; instead
Regulus is flagged as “unclassified,sibling higher” and all of the
species within Regulus are flagged as “unclassified inherited.”

The taxonomy (as of version 2.9) also contains 685 cases of taxa that
are flagged as both “unclassified inherited” and “unclassified”based
on grep unclass.*unclassified\_inherited taxonomy.tsv.

@@

Flags.
The Open Tree’s reference taxonomy is produced by a tool called
smasher that notices hints that a taxon is incertae sedis and labels
that taxon with one of five flagssee https://github.com/OpenTreeOfLife
/reference-taxonomy/wiki /Taxon-flags.This document is not concerned
with the flagging system per se, so “incertae sedis” will be used here
to refer to all of the flags that denote taxa with uncertain
placement..
While a taxonomist may intend to use incertae sedis to
indicate a limited number of possible positions for a taxon to go in
the taxonomy, we do not retain any such details.
Container nodes.
Some taxonomies have nodes with names like “Incertae
sedis (Bacteria)”.
Such nodes are called containers.
They indicate
that each child node is an incertae sedis child of the container
node's parent.
The OTT retains container nodes with the was_container
flag, but seems to have already moved all of the container's children
to the container's parent.
}

\subsection{Placement causes broken taxa}
\bdrcomment{I duplicated section \ref{conflict} to here, since we might want
  to discuss Figure \ref{fig:Handling-incertae-sedis} in the intro to answer
  why anybody would care.  Or not.  But note that the section on the ``naive approach''
  asks basically this question.}

Synthesis with \emph{incertae sedis} taxa has the potential to resolve
uncertain taxon placements using information from phylogenies.
Since
phylogenies are ranked higher than taxonomy, they may place members of
incertae sedis clades as they wish even when we do not take into
account the semantics of incertae sedis taxa.
However, without the
incertae sedis semantics described above, placing a taxon $A$ within a
taxon $B$ results in conflict with taxon $B$ in the taxonomy.
This has
three main effects: \begin{enumerate}
    \item The edge leading to clade $B$ is lost.
    \item Any taxonomy-only children of $B$ are placed at the
        MRCA of the phylogenetically informed taxa, instead of at node $B$.
    \item The name for taxon $B$ is lost.
    \item We do not record $A$ as being placed within $B$.
\end{enumerate} In order to avoid a situation
where input phylogenies conflict with a large number of taxa when
\emph{incertae sedis} taxa are placed within them, we have previously
filtered \emph{incertae sedis} taxa from the taxonomy when
constructing all prior synthesis trees.

@@

\paragraph{(A first semantics) Naive interpretation: ignoring the annotation}

The most straightforward approach for increasing the taxonomic
completeness of the supertree is simply to no longer prune taxa from
$\taxonomy$ because they are flagged as \emph{incertae sedis}, and
then run the rest of the pipeline of \citet{redelings2017supertree}
unaltered.
However this approach results in a less resolved tree
because taxonomic vagueness (in the form of the \emph{incertae sedis}
annotation) is interpreted as an unwarranted statement that the
\emph{incertae sedis} taxa are disallowed from being members of their
sibling taxa.
In essence, this interpretation would overinterpret the taxonomic
    hierarchy by treating all of the groupings as statements about
    full rooted splits, when some should be viewed as partial splits.

% \mthcomment{ Do we have some numbers on this (presumably from just
%running the only propinquity with a different set of prune flags)? }

% \bdrcomment{ I think we would currently lose about 10 taxa }

\bdrcomment{The reason why we don't do the naive thing is currently in section \ref{conflict}}
\paragraph{(Split-based semantics) Information content of the \emph{incertae sedis} annotation}

@@

When a clade is marked as \emph{incertae sedis}, we need to consider
two cases.
In the first case, the clade may be placed within a sister
clade intact (Figure \ref{fig:Handling-incertae-sedis}).
However, if
the clade is not monophyletic in the synthesis tree, then we allow the
members of the clade to be placed separately (Figure \ref{fig:Broken-incertae-sedis}).

\begin{figure}
\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree1-shaded}

\caption{\label{fig:Handling-incertae-sedis}Handling \emph{incertae sedis
}taxa recovers additional edges and taxon names.
Supertree construction
on input tree (a) and taxonomy tree (b) with \emph{incertae sedis}
clade $C$ leads to synthesis tree (c).
However, supertree construction
with \emph{incertae sedis} handling constructs tree (d), which recovers
taxon name $AB$ and $A$, as well as the edge to clade $AB$.
Taxon
names and edges that are conditional on handling \emph{incertae sedis}
are shaded grey.
}

\end{figure}
\begin{figure}
\hfill{}\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree1-shaded}

>>>>

The problem is becoming more acute because NCBI is putting a larger
number of taxa into groups that are marked as ``classified''. or
example, when OTT 2.9 was created NCBI's clasification of the bird
family Sylvidae included a group five genera that were placed within
``[http://www.ncbi.nlm.nih.gov/Taxonomy/Brows er/wwwtax.cgi?mode=Undef&
id=36270&lvl=3&lin=f&keep=1&srchmode=1&unlock||unclassi fied
Sylviidae].'' This includes the genus Regulus.
Thus in OTT 2.9 Regulus
(ott ID = 3599326) is placed inside the family Silvidae (OTT ID =
259942).
The ``unclassified Sylviidae'' does not appear in OTT; instead
Regulus is flagged as ``unclassified,sibling higher'' and all of the
species within Regulus are flagged as ``unclassified inherited.''

The taxonomy (as of version 2.9) also contains 685 cases of taxa that
are flagged as both ``unclassified inherited'' and ``unclassified''based
on grep unclass.*unclassified\_inherited taxonomy.tsv.

@@

Flags.
The Open Tree's reference taxonomy is produced by a tool called
smasher that notices hints that a taxon is incertae sedis and labels
that taxon with one of five flagssee https://github.com/OpenTreeOfLife
/reference-taxonomy/wiki /Taxon-flags.This document is not concerned
with the flagging system per se, so ``incertae sedis'' will be used here
to refer to all of the flags that denote taxa with uncertain
placement..
While a taxonomist may intend to use incertae sedis to
indicate a limited number of possible positions for a taxon to go in
the taxonomy, we do not retain any such details.
Container nodes.
Some taxonomies have nodes with names like ``Incertae
sedis (Bacteria)''.
Such nodes are called containers.
They indicate
that each child node is an incertae sedis child of the container
node's parent.
The OTT retains container nodes with the was_container
flag, but seems to have already moved all of the container's children
to the container's parent.
}
\begin{figure}
\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree1-shaded}

\caption{\label{fig:Handling-incertae-sedis}Handling \emph{incertae sedis
}taxa recovers additional edges and taxon names.
Supertree construction on input tree (a) and taxonomy tree (b) with \emph{incertae sedis}
clade $C$ leads to synthesis tree (c) if \emph{incertae sedis} annotations are
ignored, but to synthesis tree (d) if \emph{incertae sedis} annotations are respected.
The tree in (d) recovers taxon names $AB$ and $A$, as well as the edge to clade $AB$
that groups $A$ and $B$. Taxon names and edges that are conditional on handling
\emph{incertae sedis}
are shaded grey.
}

\subsection{A naive semantics: consequences of ignoring the annotation}

The most straightforward approach for increasing the taxonomic completeness of
the supertree is simply to no longer prune taxa from $\taxonomy$ because they are flagged
as \emph{incertae sedis}, and then run the rest of the pipeline of
\citet{redelings2017supertree} unaltered.  Before we propose a semantics of
\emph{incertae sedis}, it is worth pointing out what this naive semantics is lacking.

First, this approach results in a less-resolved tree because taxonomic
vagueness (in the form of the \emph{incertae sedis} annotation) is interpreted as
an unwarranted statement that the \emph{incertae sedis} taxa are disallowed from
being members of their sibling taxa.  When an incertae sedis taxon is placed within
a sibling taxon by input phylogenies, this is taken as a rejection of the monophyly
of the sibling taxon, which is then removed from the synthesis tree.  For
example, in the synthesis tree in Fig. \ref{fig:Handling-incertae-sedis}c, the
placement of taxon $C$ in group $AB$ is interpreted as a rejection of the
monophyly of $AB$. The edge leading to $AB$ is therefore not incorporated into
the synthesis tree.  In contrast, the synthesis tree in Fig.
\ref{fig:Handling-incertae-sedis}d that does not reject the monophyly of $AB$
retains the grouping of $A$ and $B$.

Second, this approach results in the removal of clade names for all clades that
have an \emph{incertae sedis} taxon place within them.  For example, in Fig.
\ref{fig:Handling-incertae-sedis}c the names for the clades $A$ and $AB$ are lost.

Third, this approach cannot use information from phylogenies to resolve the
location of uncertain taxa.  For example, the synthesis tree in
\ref{fig:Handling-incertae-sedis}c has lost the label for $A$, and so does not
allow us to place $C$ within $A$.  In contrast, the synthesis tree in Fig.
\ref{fig:Handling-incertae-sedis}d that takes account of \emph{incertae sedis}
annotations and retains the label for $A$.  This allows us to infer that
\emph{incertae sedis} taxon $C$ has been placed within $A$, resolving
the uncertainty.

@@

\paragraph{(Split-based semantics) Information content of the \emph{incertae sedis} annotation}

@@

When a clade is marked as \emph{incertae sedis}, we need to consider
two cases.
In the first case, the clade may be placed within a sister
clade intact (Figure \ref{fig:Handling-incertae-sedis}).
However, if
the clade is not monophyletic in the synthesis tree, then we allow the
members of the clade to be placed separately (Figure \ref{fig:Broken-incertae-sedis}).
\begin{figure}
\hfill{}\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree1-shaded}
